\documentclass{article}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{bookmark}
\usepackage{float} 
\usepackage{graphicx}
\usepackage{makeidx}
\usepackage[letterpaper, total={7.5in, 10in}]{geometry}

\makeindex

\setlength{\parindent}{0pt}
\begin{document}
\title{De Bellum Gallico IV}
\author{Elias Xu}
\date{\today}
\maketitle

\setlength{\parindent}{0pt}

\section*{24}

\subsection*{1}
At barbarī, cōnsiliō Rōmānōrum cognitō, praemissō equitātū et ēssedariīs, quō plērumque genere in proeliīs ūtī consuērunt, reliquīs copiīs subsecūtī nostrōs nāvibus ēgredī prohibēbant.  \\
\textbf{But the barbarians, with plan of Romans having been known, with horsemen and charioteers, whom they are accustomed to to use in battle, were preventing our men from disembarking from ships.}

\subsection*{2}
Erat ob hās causās summa difficultās, quod navēs propter magnitūdinem nisi in altō cōnstituī nōn poterant, mīlitibus autem, ignōtīs locīs, impedītīs manibus, magnō et gravī onere armōrum oppressīs simul et dē nāvibus desiliendum et in fluctibus consistendum et cum hostibus erat pugnandum,  \\
\textbf{It was because this occurrence the greatest difficulty, because, on on account of their magnitude, the ships were not able to be stopped except in deep [sea], the soldiers, however, in unknown place, with impeded hands having been overcome by great and heavy burdens of armor and had to step off from ships and stand still in waves and fight with the enemy,}

\subsection*{3}
cum illī aut ex āridō aut paulum in aquam progressī omnibus membrīs expedītīs, nōtissimīs locīs, audacter tēla coicerent et equōs insuēfactōs incitārent.  \\
\textbf{who to that were either dry or small in water, advancing having been unencumbered, to most well know place, boldly threw together darts, incited accustomed horses.}

\subsection*{4}
Quibus rēbus nostrī perterritī atque huius omnīnō generis pugnae imperītī, nōn eādem alacritāte ac studiō quō in pedestribus ūtī proeliīs consuerant utēbantur. \\
\textbf{By these things out men having been frightened greatly and in these all unskilled of these things were using not with same vigur and zeal which then accustomed to use in foot battle. }


\section*{25}
\subsection*{1}
Quod ubi Caesar animadvertit, nāvēs longās, quārum et speciēs erat barbarīs inūsitātior et motus ad usum expedītior, paulum removērī ab onerāriīs nāvibus et rēmīs incitārī et ad latus apertum hostium constituī atque inde fundīs, sagittīs, tormentīs hostēs propellī ac submovērī iussit; quae rēs magnō usuī nostrīs fuit. \\
\textbf{Because where Caesar had turned his mind, he ordered long ships, and of whose form was more uncommon to barbarians and could be moved towards use more unimpeded, some to be removed from ships of burdens and to be excited by oars and to be stopped still on open flank of enemy and from that place with slings, arrows, and missle launchers, to push away and move enemy, of which things were for our men a great use.}

\subsection*{2}
Nam et nāvium figūrā et rēmōrum mōtū et inusitātō genere tormentōrum permōtī barbarī constitērunt ac paulum modo pedem rettulērunt.\\
\textbf{For both by form of ships and movement of oars and uncommon kind of missile launcher from an unaccustomed origin disturbed barbarians and a little by this matter brought back feet.}

\subsection*{3}
Atque nostrīs mīlitibus cunctantibus, maximē propter altitūdinem maris, quī X legiōnis aquilam ferēbat, contestātus deōs ut ea rēs legiōnī feliciter ēvenīret, ‘Dēsilite,’ inquit, ‘militēs, nisi vultis aquilam hostibus prōdere; ego certē meum rēī pūblicae atque imperātōrī officium praestiterō.’\\
\textbf{And with our soldiers delaying, especially on account of the height of the sea, someone who bore legion X eagle, having bore witness to gods so that happy circumstance for legion will turn out well, he said, "Jump down, soldiers, unless you want to betray standard to enemies. I certainly will have placed aove myself duty to public things and to commander." }

\subsection*{4}
Hoc cum vōce magnā dīxisset, sē ex nāvī prōiēcit atque in hostēs aquilam ferre coepit. \\
\textbf{This he said with great voice, he threw himself forward from ships and began to bear standard to enemy.}

\subsection*{5}
Tum nostrī cohortātī inter sē, nē tantum dēdecus admitterētur, universī ex nāvī dēsiluērunt. \\
\textbf{Then our men having been exhorted among themselves, lest so great shame is allowed, universally, they jumped down from ships.}

\subsection*{6}
Hōs item ex proximīs primī nāvibus cum conspēxissent, subsecutī hostibus appropinquāverunt.\\
\textbf{Likewise when from the neighboring ships they beheld this, following they approached having pursued enemies.}







\section*{26}

\subsection*{1}
Pugnātum est ab utrīsque ācriter. Nostrī tamen, quod neque ordinēs servāre neque firmiter insistere neque signa subsequī poterant atque alius aliā ex nāvī quibuscumque signīs occurrerat sē aggregābat, magnopere perturbābantur; \\
\textbf{Bitterly it is having been fought by all sides. Our men nevertheless, because there was unable neither to preserve ranks nor firmly follow signs and one from one ship aggregates himself to whatever signs occurs from another ship, were greatly disturbed}

\subsection*{2}
hostēs vērō, nōtīs omnibus vadīs, ubi ex lītore aliquōs singulārēs ex nāvī ēgredientēs conspēxerant, incitātīs equīs impedītōs adoriēbantur, \\
\textbf{Enemies truly with all shallowness having been known, beheld where some of men fleeing from ships, attacked our men impeded with calvary having been aroused}

\subsection*{3}
plūrēs paucōs circumsistēbant, aliī ab latere apertō in universōs tēla coniciēbant.\\
\textbf{many surrounded by dew, others were throwing spears all together from wide flanks}

\subsection*{4}
Quod cum animadvertisset Caesar, scaphās longārum nāvium, item speculātōria navigia mīlitibus complērī iussit et, quōs laborantēs conspēxerat, hīs subsidia submittēbat. \\
\textbf{Because wen Caesar had turned his mind, he ordered skiff of longer boats likewise scout ships to be filled with soldiers, where [who was] laboring he beheld to them he was sending up aid}

\subsection*{5}
Nostrī, simul in āridō constitērunt, suīs omnibus consecūtīs, in hostēs impetum fecērunt atque eōs in fūgam dedērunt; neque longius prosequī potuērunt, quod equitēs cursum tenēre atque insulam capere non potuerant. Hoc ūnum ad pristīnam fortūnam Caesarī defuit.\\
\textbf{Our men, gathered at the same time in dry [land] with all of their own following after, made an attack in enemies and gave them into flight, and not were able to pursue longer, because they were not able to hold calvary's course and seize island, those one lacked to pristine fortune of Caesar.}






\section*{27}
\subsection*{1}
Hostēs proeliō superātī, simul atque sē ex fūgā recēpērunt, statim ad Caesarem lēgātōs dē pace mīsērunt; obsidēs datūrōs quaeque erāsset sēsē factūrōs pollicitī sunt. \\
\textbf{Enemies having been overcome by the battle, as sone as they had recovered themselves from the battle immediately they sent legates to Caesar about peace, they promised they will give hostages and do what things he [Caesar] had ordered.}

\subsection*{2}
Ūnā cum hīs legātīs Commius Atrēbās vēnit, quem suprā demonstrāveram ā Caesare in Britanniam praemissum. \\
\textbf{Together with this legate one C. Atrebas who I above demonstrated in Britannica hav ing been sent forward by Caesar,}

\subsection*{3}
Hunc illī ē nāvī ēgressum, cum ad eōs ōrātōris mōdō Caesaris mandāta deferret, comprehenderant atque in vincula coiēcerant; \\
\textbf{They had arrested this man having exiting a ship, when to them Commius carried in a manner of a speaker Caesar's orders, they seized and and threw him in chains.}

\subsection*{4}
tum proeliō factō remīsērunt. In petendā pace eius rēī culpam in multitūdinem coiēcērunt et propter imprudentiam ut ignōscerētur petivērunt.\\
\textbf{They then sent him back with battle having been made. In seeking pease they threw together blame of affair on multitude, and on account of lack of foresight they sought so they would be pardoned.}

\subsection*{5}
Caesar questus quod, cum ultrō in continentem lēgātīs missīs pacem ab sē petissent, bellum sine causā intulissent, ignōscere <sē> rudentiae dixit obsidēsque imperāvit; 
\textbf{Caesar complaining because, although they had sought peace from him, with legates that had been sent into the continent they made wat without cause, he would forgive their impudence and he ordered for their [hostages] .}

\subsection{5}
quōrum illī partem statim dedērunt, partem ex longinquiōribus locīs arcessitam paucīs diēbus sēsē datūrōs dīxērunt. \\
\textbf{a part of whom the enemies gave immediately another they said they would give in a few days having been summoned from distant places.}

\subsection*{6}
Intereā suōs remigrāre in agrōs iussērunt, principēsque undique convenīre et sē civitātēsque suās Caesarī commendāre coepērunt.\\
\textbf{Meanwhile they ordered their people to return to the field and assemble from all quarters and began to surrender themselves and their own states to Caesar.}

\section*{28 (N.T)}
\subsection*{1}
Hīs rēbus pāce confirmātā, post diem quartum quam est in Britānniam ventum nāvēs XVIII, dē quibus suprā demonstrātum est, quae equitēs sustulerant, ex superiōre portū lēnī ventō solvērunt. \\
\textbf{With peace having been confirmed by these affairs, after the fourth day there was an arrival into Britannia 18 ships about which it has been mentioned above, which had taken the calvary aboard, loosened out of the further port with light breeze. }

\subsection*{1}
Quae cum appropinquārent Britānniae et ex castrīs vidērentur, tanta tempestās subitō coörta est ut nūlla eārum cursum tenēre posset, sed ae eōdem unde erant profectae referrentur, aliae ad inferiōrem partem insulae, quae est prōpius sōlis occāsum, magnō suī cum perīculō deicerentur; \\
\textbf{When they approached Britannia and were seen from the camps, a sudden so great a storm rose with the result that none of them were able to hold their course, but rather some were carried back from where they had left, other were thrown to a lower part of the island which is closer to the setting of the sun with their own great danger.}

\subsection*{1}
quae tamen, ancorīs iactīs, cum fluctibus complērentur, necessāriō adversā nocte in altum provectae continentem petiērunt.\\
\textbf{Nevertheless, with the anchors having been thrown because they were being filled with waves, by necessity on the opposing night they sought the continent having been carried into the deep.}

\section*{29}

\subsection*{1} 
Eādem nocte accidit ut esset luna plēna, quī diēs maritimōs aestūs maximōs in Ōceanō efficere consuēvit, nostrīsque id erat incognītum. \\
\textbf{Likewise it befell in night that moon was fully, which is accustomed that maritime days swell greatest tide in ocean, and it was unknown to our men.}

\subsection*{2} 
Ita ūnō tempore et longās nāvēs, quibus Caesar exercitum transportandum curāverat, quāsque in aridum subdūxerat, aestus complēverat, et onerāriās, quae ad ancorās erant deligātae, tempestās afflictābat, neque ūlla nostrīs facultās aut administrandī aut auxiliandī dabātur.\\
\textbf{Already with one time and long ships, and transport ships, with which Caesar carried for transporting soldiers, and which he led in dry [land], tide filled up, which having been tied down with anchors, tempest was afflicting, and not opportunity was given or faculty for management or help.}

\subsection*{3} 
Complūribus nāvibus fractīs, reliquae cum essent fūnibus, ancorīs reliquīsque armamentīs amissīs ad navigandum inūtilēs, magna, id quod necesse erat accidere, tōtīus exercitūs perturbātio facta est. \\
\textbf{With good number of ships having been broken, remaining with ropes, anchors, and remaining armaments having been let loose, were useless for navigating well, disturbing all that which was necessary to occur,  }

\subsection*{4} 
Neque enim nāvēs erant aliae quibus reportārī possent, et omnia dēerant quae ad reficiendās nāvēs erant usuī, et, quod omnibus constābat hiemārī in Galliā oportere, frumentum in hīs lōcīs in hiemem provīsum nōn erat.\\
\textbf{And not indeed ships were of other by which soldiers are able to be carried back, and all was lacking with which was for repairing of ship, and the grain, which was evident to many to be appropriate for spending the winter in Gaul had not been provided in these places for winter. }


\section*{30}

\subsection*{1}
Quibus rēbus cognitīs, principēs Britānniae, quī post proelium ad Caesarem convēnerant, inter sē collocutī, cum equitēs et nāvēs et frumentum Rōmānīs dēesse intellegerent et paucitātem mīlitum ex castrōrum exiguitāte cognōscerent, quae hōc erant etiam angustiora quod sine impedimentīs Caesar legiōnēs transportāverat.\\
\textbf{With all of these things, with all having been recognized, Chiefs of Britannia, with all having been gathered after the battle to Caesar, because they understand calvary, ships, and grain were lacking to the Romans, were recognizing the paucity of soldiers from smallness or military camp, which were more narrow because of this, because which Caesar had transported there legions without a bunch of baggage. }

\subsection*{2}
optimum factū esse dūxērunt rebelliōne factā frūmentō commeātūque nostrōs prohibēre et rem in hiemem prodūcere, quod eīs superātīs aut redītū interclusīs nēminem posteā bellī inferendī causā in Britanniam transitūrum cōnfidēbant. \\
\textbf{They believed that that it was best to do, with rebellion having been made, prohibited from grain and going back and forth and produce stuff to winter, because with the men having been conquered or having been cut off from return, they were trusting that no one would travel into Britain for the sake of starting war. }

\subsection*{3}
Itaque rursus coniurātiōne factā paulātim ex castrīs discedere et suōs clam ex agrīs dēdūcere coepērunt.\\
\textbf{And so with a conspiracy having been done little by little departed from camps and led me secretly from field.}


\section*{31 (N.T.?)}
\subsection*{1}
At Caesar, etsī nōndum eōrum cōnsilia cognōverat, tamen et ex ēventū nāvium suārum et ex eō quod obsidēs dare intermīserant fore id quod accidit suspicābātur. \\
\textbf{But Caesar, although he had not yet recognized the plan, nonetheless he was suspicious that the result of his own ships, and out of that fact, that enemies had stopped to give hostages, hw was suspecting that which will occur which has happened. }

\subsection*{2}
Itaque ad omnēs casūs subsidia comparābat. Nam et frūmentum ex agrīs cotidiē in castra cōnferēbat et, quae gravissimē afflictae erant nāvēs, eārum materiā atque aere ad reliquās reficiendās ūtēbātur et quae ad eās rēs erant usuī ex continentī comportārī iubēbat. \\
\textbf{And therefore he was preparing help for all types of calamity for he was daily conferring grain from the field into the camps, and was using bronze and material of the bronze of the ships having been damaged to repairing remaining ships was ordering [that] which were a use for these affairs  to be carried from continent.}

\subsection*{3}
Itaque, cum summō studiō ā mīlitibus administrāretur, XII nāvibus āmissīs, reliquīs ut navigārī satis commodē posset effēcit.\\
\textbf{And so, it was being administered with soldiers with highest zeal, although 12 ships were lost, he made for the rest [of the ships] to be able to sail away safely.}


\section*{32}
\subsection*{1}
Dum ea geruntur, legiōne ex consuetūdine ūnā frumentātum missā, quae appellābātur VII., neque ūllā ad id tempus bellī suspiciōne interpōsitā, cum pars hominum in agrīs remanēret, pars etiam in castra ventitāret, eī quī prō portīs castrōrum in statiōne erant Caesarī nuntiāvērunt pulverem maiōrem quam consuetūdō ferret in eā parte vidērī quam in partem legiō iter fēcisset.\\
\textbf{While it was borne, with the legion from custom one having bee sent for the purpose of gathering grain, which was called VII, and with suspicion of anything to it time of war having been alleged, with parts of men remaining in fields, part also came intocamp, and who were before gates of the camp in the station of Caesar announced that greater dirt than that was accustomed carried in each part were a legion would make in a journey. }

\subsection*{2}
Caesar id, quod erat, suspicātus, aliquid nōvī ā barbarīs inītum cōnsiliī, cohortēs quae in statiōne erant sēcum in eam partem proficīscī, ex reliquīs duās in statiōnem succēdere, reliquās armārī et confestim sēsē subsequī iussit.\\
\textbf{This Caesar, because he was, suspecting, this was that something of a new plan having been begun from the barbarians, ordered cohorts which were in station with himself to advance to that part, two from remaining to go up, remaining to be armed and immediately follow himself.}

\subsection*{3}
Cum paulō longius ā castrīs processisset, suōs ab hostibus premī atque aegrē sustinēre et, cōnfertā legiōne, ex omnibus partibus tēla coīcī animadvertit. \\
\textbf{When a little farther it was advancing from camp, he noticed his men pressed together by the enemy, with legion having been compacted weakly sustaining and from all parts spears are thrown.}

\subsection*{4}
Nam quod omnī ex reliquīs partibus demessō frūmentō pars ūna erat reliqua, suspicātī hostēs hūc nostrōs esse ventūrōs noctū in silvīs delituerant; \\
\textbf{For because with all grain having been reaped from remaining parts enemies suspecting that our men would go into at night, hid in forest }

\subsection*{5}
tum dispersōs depōsitīs armīs in metendō occupātōs subitō adortī paucīs interfectīs reliquōs incertīs ordinibus perturbāverant, simul equitātū atque ēssedīs circumdederant.\\
\textbf{then sudeenly attacked our men with a few men having been killed disturbed our ranks and same time surrounded with horsemen and calvary.}


\section*{33}
\subsection*{1}
Genus hoc est ex ēssedīs pugnae. Prīmō per omnēs partēs perequitant et tēla coīciunt atque ipsō terrōre equōrum et strepitū rotārum ordinēs plērumque perturbant et, cum sē inter equitum turmās insinuāverunt, ex ēssedīs dēsiliunt et pedibus proeliantur. \\
\textbf{This is kind of fighting from a chariot. First they ride a horse through all parts and throw spears and with own terror of horses and with creaking of wheels disturb order of our men and when they would wind their way in among the horsemen jump off chariots and fight with feet.}

\subsection*{2}
Aurigae interim paulātim ex proeliō excēdunt atque ita currūs conlocant ut, sī illī ā multitudine hostium premantur, expedītum ad suōs receptum habeant. \\
\textbf{Meanwhile charioteers leave little by little and place chariots in such a way that if men are pressed by multitude, by enemy they would have an easy way to retreat.}

\subsection*{3}
Ita mobilitātem equitum, stabilitātem peditum in proeliīs praestant, ac tantum ūsū cotidiānō et exercitātiōne efficiunt utī in declivī ac praecipitī lōcō incitātōs equōs sustinēre et brevī moderārī ac flectere et per temonem percurrere et in iugō insistere et sē inde in currūs citissimē recipere consuerint.\\
\textbf{So they show mobility of calvary and stability of infrantryman in battle and effect thus so greatly because by means of daily use and practice with the result that they are accustomed to first sustain horses having been roused on step and declining places guide and turn them in a short time able to run through pole and to stand on yoke and take themselves back very quickly and run back easily. }


\section*{34}
\subsection*{1}
Quibus rēbus perturbātīs nostrīs novitate pugnae tempore opportunissimō Caesar auxilium tulit: namque eius adventū hostēs constitērunt, nostrī sē ex timore recepērunt. \\
\textbf{With these things having perturbed our men with newness of fighting, on time most opportune Caesar bore help, and because enemies stopped with his arrival, our men themselves retreated from fear.}

\subsection*{2}
Quō factō, ad lacessendum hostem et committendum proelium aliēnum esse tempus arbitrātus suō sē locō continuit et brevī tempore intermissō in castra legiōnēs redūxit. \\
\textbf{With which having been made, for sake of attacking enemy and for committing of battle other time having been thought of help himself in his own place, and he led with short time in camp legions.}

\subsection*{3}
Dum haec geruntur, nostrīs omnibus occupātīs quī erant in agrīs reliquī discessērunt. \\
\textbf{While this was borne, with all men having been occupied those who were in field remaining reparted.}

\subsection*{4}
Secutae sunt continuōs complurēs diēs tempestātēs, quae et nostrōs in castrīs continērent et hostem ā pugnā prohibērent.\\
\textbf{Storm followed for continuing days, of which both contained our men in camp and prohibited enemies from fighting}

\subsection*{5}
Interim barbarī nuntiōs in omnēs partēs dimīsērunt paucitātemque nostrōrum mīlitum suīs praedicāvērunt et quanta praedae faciendae atque in perpetuum suī liberandī facultās darētur, sī Rōmānōs castrīs expulissent, demonstrāvērunt. Hīs rēbus celeriter magnā multitudine peditātūs equitātūsque coactā ad castra vēnērunt.\\
\textbf{Meanwhile barbarians sent messages in all parts and proclaimed paucity of our soldiers to themselves and how much opportunity of making reward for freedom and in perpetuity of freeing themselves is giving. If they expect romand from camps, they demostrated, to theset things swiftly with a great multitude of foot soldiers and horsemen having been acting they came to campp. }

\section*{35}
\subsection*{1}
Caesar, etsī idem quod superiōribus diēbus acciderat fore vidēbat, ut, sī essent hostēs pulsī, celeritāte perīculum effugerent, tamen nactus equitēs circiter XXX, quōs Commius Atrēbās, dē quō ante dictum est, sēcum transportāverat, legiōnēs in aciē prō castrīs constituit.  \\
\textbf{Caesar, although the same befell which was seen to happen with following days so that, if enemies were driven out, they would swiftly flee danger nevertheless having obtained around 30 horsemen, who, C. Atreban, about who was said before, transported withm them, placed legions in a battle line before camps.}

\subsection*{2}
Commissō proeliō diutius nostrōrum mīlitum impetum hostēs ferre non potuērunt ac terga vertērunt.  \\
\textbf{After a battle having been commenced, for a long time enemeies were not able to bear attack by our soldiers and turned back.}

\subsection*{3}
Quōs tantō spatiō secutī quantum cursū et vīribus efficere potuērunt, complūrēs ex eīs occidērunt, deinde omnibus longē lātēque aedificiīs incensīs sē in castra recēpērunt. \\
\textbf{Who with great space following they were able to execute as much with course and strength, many from them died they took back themselves with long to all and with wide buildings having been burning in camp.}


\section*{36}
\subsection*{1}
Eōdem diē lēgātī ab hostibus missī ad Caesarem dē pace vēnērunt.\\
\textbf{Following day Legates having been sent from enemies came to Caesar concerning.}


\end{document}